---
title: "Phase 2 - Regression Analysis: Drivers of Data Center CO2 Emissions"
author: "Lakshay Mani"
date: "2026-03-30"
output: html_document
---

## Overview

This analysis identifies which factors statistically drive CO2 emissions 
from U.S. data centers, using OLS regression with three predictors:
energy consumption, grid carbon intensity, and renewable energy percentage.

**Data Sources:**

- LBNL 2024 US Data Center Energy Usage Report
- Ember (2026) via Our World in Data
- EIA Monthly Energy Review March 2026, Table 2.6

---

## Load Data

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

df <- read.csv("master_dataset.csv")
head(df)
str(df)
```

---

## Correlation Analysis

Before regression, check how each variable relates to emissions.

```{r correlation}
cor_matrix <- cor(df[, c("DC_Energy_TWh", 
                          "Grid_CarbonIntensity_gCO2perKWh",
                          "Renewable_Pct",
                          "Derived_CO2_MtCO2")])
round(cor_matrix, 3)
```

---

## OLS Regression Model

```{r regression}
model <- lm(Derived_CO2_MtCO2 ~ DC_Energy_TWh + 
                                  Grid_CarbonIntensity_gCO2perKWh + 
                                  Renewable_Pct, 
            data = df)
summary(model)
```

---

## Multicollinearity Check (VIF)

VIF values above 5 indicate high multicollinearity, above 10 are severe.
Given the high correlations between predictors, this check confirms whether
the full model coefficients should be interpreted with caution.

```{r vif_check, message=FALSE, warning=FALSE}
if (!require(car)) install.packages("car", repos="http://cran.r-project.org")
library(car)
vif(model)
```

---

## Visualize: Actual vs Fitted

```{r fitted_plot}
df$Fitted <- fitted(model)

ggplot(df, aes(x = Year)) +
  geom_line(aes(y = Derived_CO2_MtCO2, color = "Actual"), linewidth = 1) +
  geom_line(aes(y = Fitted, color = "Model Fitted"), 
            linetype = "dashed", linewidth = 1) +
  labs(title = "Regression Model: Actual vs Fitted CO2 Emissions",
       x = "Year", y = "CO2 Emissions (Mt CO2)",
       color = "Series") +
  theme_minimal()
```

---

## Individual Predictor Models

```{r simple_models}
# Model 1: Energy alone
model_simple <- lm(Derived_CO2_MtCO2 ~ DC_Energy_TWh, data = df)
summary(model_simple)

# Model 2: Carbon intensity alone
model_ci <- lm(Derived_CO2_MtCO2 ~ Grid_CarbonIntensity_gCO2perKWh, data = df)
summary(model_ci)

# Model 3: Renewables alone
model_ren <- lm(Derived_CO2_MtCO2 ~ Renewable_Pct, data = df)
summary(model_ren)

# Compare R-squared across all three
cat("R-squared comparison:\n")
cat("Energy alone:           ", round(summary(model_simple)$r.squared, 4), "\n")
cat("Carbon intensity alone: ", round(summary(model_ci)$r.squared, 4), "\n")
cat("Renewables alone:       ", round(summary(model_ren)$r.squared, 4), "\n")
```

---

## Phase 2 Findings

OLS regression confirms **DC energy consumption** is the dominant 
driver of emissions (R² = 0.9959, p < 0.001). Every additional 
1 TWh of data center energy adds **0.33 Mt CO2**.

Carbon intensity (R² = 0.785) and renewable percentage (R² = 0.825) 
are statistically significant when isolated but secondary to energy volume.

Note: High VIF values are expected given the strong correlation between 
energy growth and renewable adoption over time. Individual predictor 
models are therefore the cleaner analytical lens.

**Implication:** LBNL's emissions underestimate is not from 
mispredicting energy growth — it's from underestimating the carbon 
intensity of the specific grids powering data centers.
